Hydra battles in Coq

### 2020-09-15 doc/chapter-ordinals.tex
minor modifications

\subsection{Countable ordinals}

@@

\subsection{Ordinal notations}

@@

Let $\alpha$ be some (countable) ordinal; 
in \coq{} terms, we call ordinal notation for $\alpha$ a structure composed 
of:
\begin{itemize}
\item A data type $A$ for representing all ordinals strictly below $\alpha$,
\item A well founded order $<$ on $A$, 
\item A Function for comparing two ordinals, which implies that the reflexive closure of $<$ is a \emph{total order}.
\end{itemize}

@@

\subsection{Ordered types}

@@

\begin{remark}
The infix notations \texttt{o<} and \texttt{o<=} were defined in order to make apparent the distinction between the various notation scopes that may co-exist in a statement. In the informal and semi-formal statements, we still use $<$ and $\leq$.
\end{remark}

\section{Instances of notation systems}

@@

In order to build an instance of \texttt{ON}, we have to define a comparison function.

@@

We can now define abbreviations. For instance, the finite ordinals will be represented by the terms built with  the constructor \texttt{inl}, and the first infinite ordinal $\omega$ by the term \texttt{(inr 0)}.

@@

\begin{itemize}
\item A \emph{least} element is a minorant (in the large sense) of the full set  on $A$,
\item $y$ is a \emph{successor} of $x$ if $x<y$ and there is no element between $x$ and $y$,
\item $x$ is a \emph{limit} if $x$ is not a least element, and for any $y$ such that $yo<x$,
 there exists some $z$ such that $y<z<x$.
\end{itemize}

The following definitions are in Library\href{../src/html/hydras.Prelude.MoreOrders.html}{Prelude.MoreOrders}.

@@

@@

Using thedefinitions above, we can prove the following lemma:

@@

Notation "'omega'" := (1,0) : o2_scope.
\end{Coqsrc}

@@

The following class specifies a strict inclusion of segments. The notation \texttt{OA} describes a segment $[0,alpha($, and \texttt{OB} is a larger segment (which contains a notation for $\alpha$, whilst $\alpha$ is not represented by any term of type $A$). We require also  that the comparison functions of the two notation systems are compatible.

@@

\begin{Coqsrc}
Class  SubON 
       `(OA : @ON A ltA  compareA)
       `(OB : @ON B ltB  compareB)
       (alpha :  B)
       (iota : A -> B):=
  {
  subseg_compare :forall x y : A,  compareB (iota x) (iota y) =
                                 compareA x y;
  subseg_incl : forall x, ltB (iota x) alpha;
  subseg_onto : forall y, ltB y alpha  -> exists x:A, iota x = y}.
\end{Coqsrc}

@@

\subsection{Comparing an ordinal notation with Schütte's model}

@@

We are now able to build an instance of \texttt{Subsegment}.

>>>>

Two kinds of representation of ordinals are defined herein.

\begin{itemize}
\item A ``mathematical'' representation of the set of countable ordinal numbers, afer Kurt Schütte~\cite{schutte}. This representation uses several (hopefully harmess) axioms. It is used as a reference for proving the correctness of ordinal notations.
\item A family of \emph{ordinal notations}, \emph{i.e.} data types used to represent segments $[0,\mu)$, where $\mu$ is some countable ordinal. Each ordinal notation is defined inside the Calculus of Inductive Constructions (without axioms). Many functions are defined, allowing proofs by computation. Note that proofs of 
correctness of a given ordinal notation with respect to Schütte's model use axioms.
Please execute the \texttt{Print Assumptions} command in case of doubt.
\end{itemize}

\section{Countable ordinals}

@@

\section{Ordinal notations}

@@

Let $\alpha$ be some (countable) ordinal; 
in \coq{} terms, we call ordinal notation for $\alpha$ a structure composed 
of:
\begin{itemize}
\item A data type $A$ for representing all ordinals strictly below $\alpha$,
\item A well founded order $<$ on $A$, 
\item A function for comparing two ordinals, whose correctnes implies that the reflexive closure of $<$ is a \emph{total order}.
\end{itemize}

@@

\subsubsection*{Ordered types}

@@

\begin{remark}
The infix notations \texttt{o<} and \texttt{o<=} were defined in order to make apparent the distinction between the various notation scopes that may co-exist in a same statement. So the infix \texttt{<} and \texttt{<=} are reserved to the natural numbers. In the mathematical formulas, we still use $<$ and $\leq$ for comparing ordinals.
\end{remark}

\section{Examples of ordinal notations}

@@

Before building an instance of \texttt{ON}, we have to define a comparison function.

@@

We can now define abbreviations. For instance, the finite ordinals are represented by terms built with  the constructor \texttt{inl}, and the first infinite ordinal $\omega$ by the term \texttt{(inr 0)}.

@@

\begin{itemize}
\item A \emph{least} element is a minorant (in the large sense) of the full set  on $A$,
\item $y$ is a \emph{successor} of $x$ if $x<y$ and there is no element between $x$ and $y$. We will also say that $x$ is a \emph{predecessor} of $y$.
\item $x$ is a \emph{limit} if $x$ is not a least element, and for any $y$ such that $yo<x$,
 there exists some $z$ such that $y<z<x$.
\end{itemize}

The following definitions are in Library \href{../src/html/hydras.Prelude.MoreOrders.html}{Prelude.MoreOrders}.

@@

\index{Exercises}
\begin{exercise}
Prove, that, in any ordinal notation system, that every ordinal has at most one predecessor, and at most one successor. 
\end{exercise}

@@

Using the definitions above, we can prove the following lemma:

@@

Notation "'omega'" := (1,0) : o2_scope.

Definition limitb (alpha : t) := 
 match alpha with
 |  (S _, 0) => true
 | _ => false
end.
\end{Coqsrc}

In order to build an ordinal notation out the type \texttt{t}, we have to define a comparison function, and prove that the order \texttt{lt} is well-founded.

\begin{Coqsrc}
Definition compare (alpha beta: t) : comparison :=
  match Nat.compare (fst alpha) (fst beta) with
    Eq => Nat.compare (snd alpha) (snd beta)
  | c => c
  end.

Lemma compare_correct alpha beta :
    CompareSpec (alpha = beta) (lt alpha beta) (lt beta alpha)
                (compare alpha beta).

Instance Omega2 : ON  lt compare.
\end{Coqsrc}

@@

The following class specifies a strict inclusion of segments. The notation \texttt{OA} describes a segment $[0,\alpha($, and \texttt{OB} is a larger segment (which contains a notation for $\alpha$, whilst $\alpha$ is not represented in \texttt{OA}). We require also  that the comparison functions of the two notation systems are compatible.

@@

\begin{Coqsrc}
Class  SubON 
       `(OA : @ON A ltA  compareA)
       `(OB : @ON B ltB  compareB)
       (alpha :  B)
       (iota : A -> B):=
  {
  subon_compare :forall x y : A,  compareB (iota x) (iota y) =
                                 compareA x y;
  subon_incl : forall x, ltB (iota x) alpha;
  subon_onto : forall y, ltB y alpha  -> exists x:A, iota x = y}.
\end{Coqsrc}

@@

\section{Comparing an ordinal notation with Schütte's model}

@@

We are now able to build an instance of \texttt{SubON}.
